---
title: "MCL_Testdatensatz"
author: "Heona"
date: "`r Sys.Date()`"
output: 
  html_document:
    css: styles.css
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: darkly
    highlight: tango
    number_sections: false
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
  options:
    error: TRUE
    warning: TRUE
    message: TRUE
---
Für md output: 
  gfm : 
    code-fold: false
editor: visual
date: "`r Sys.Date()`"
knitr:
  opts_chunk:
    message: false
    warning: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(janitor)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(cowplot)
library(randomcoloR)
library(ComplexHeatmap)
library(circlize)
library(tibble)
library(psych)
library(pcaMethods)
library(scrime)
```

# MCL testdata

## Import the data

First I need to import the RDS file to R-Studio.

I assign it to mcl_testdata and standardize the format of column names by clean_names() from the janitor package.

--\> removes all weird white spaces/spaces/special characters/etc.
and replaces with underscores(\_)

```{r data import}
getwd()
setwd("/Users/hlee/Desktop/Doktorarbeit")
mcl_testdata_original <- readRDS("20230727_MCL_testdata.rds") %>%
  janitor::clean_names()
```

```{r dimensions of original raw data}
dim(mcl_testdata_original)
```

The original dataset consists of 7211 rows and 262 columns.

I only need uniprot_i\_ds, gene_names corrected reporter intensity information.

--\> I only select the relevant columns

--\> I rename them to uniprot_id and gene_names and remove "reporter_intensity_corrected" from the column names (except for in uniprot_id and gene_names)

```{r selecting and renaming data}
mcl_testdata <- mcl_testdata_original %>%
  dplyr::select(uniprot_i_ds, gene_names, contains("reporter_intensity_corrected")) %>%
  dplyr::rename("uniprot_id" = "uniprot_i_ds" ) %>%
  dplyr::rename("gene_id" = "gene_names") %>%
  dplyr::rename_with(.cols = -c("uniprot_id", "gene_id"), .fn = ~ stringr::str_remove(string = ., "reporter_intensity_corrected_"))

dim(mcl_testdata)
```

The transformed dataset mcl_data contains 7211 rows and 222 columns.

## Quick data checkup

### Missing values

Check for number of missing values R likes to work with NA as missing values, but in out dataset missing values are NaN.

```{r missing values}
#convert NaN to NA
#this code replaces "NaN" with NA of type integer:
mcl_testdata[mcl_testdata == "NaN"] <- NA_integer_

#convert Zero values to NA
#this code replaces 0 with NA of type integer except for column 1 and 2
mcl_testdata[,-c(1:2)][mcl_testdata[,-c(1:2)] == "0"] <- NA_integer_

#remove rows with only NA
mcl_testdata <- mcl_testdata %>%
  filter(rowSums(is.na(mcl_testdata[,-c(1:2)])) != ncol(mcl_testdata[,-c(1:2)]))

print(paste("Total number of missing values: ", sum(is.na(mcl_testdata))))

dim(mcl_testdata)
```

The dataset now contains 7007 rows and 222 columns.

### Duplicates

check for duplicated features and samples:

```{r duplicates}
print(paste("Number of duplicated features: ", sum(duplicated(mcl_testdata$uniprot_id))))
print(paste("Number of duplicated samples: ", sum(duplicated(colnames(mcl_testdata[,-c(1:2)])))))
```

### Combine replicates & reshape dataframe from wide to long

with pivot_longer() function

--\> the columns are transformed into 2 new columns: "Prot_id1" (previously the column names that contained "r1" = replicate 1) and "count_R1" (corresponding to the values in those columns)

--\> the same thing with Prot_id2 and count_R2 (for replicate 2)

--\> then: a new column containing the mean of count_R1 and count_R2 is added, ignoring the NAs in the mean calculation

```{r attempt 1 to combine replicates, eval = FALSE}
#combine both replicates 
mcl_testdata_long <- cbind(
  mcl_testdata %>%
    dplyr::select(uniprot_id, contains("r1")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "Prot_id1",
                        values_to = "count_R1"),
  mcl_testdata %>%
    dplyr::select(uniprot_id, contains("r2")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "Prot_id2",
                        values_to = "count_R2") %>%
    dplyr::select(-uniprot_id)
    
) %>%
  rowwise() %>%
  dplyr::mutate(avg_count = mean(c_across(cols = c(count_R1, count_R2)), na.rm = TRUE))
```

somehow, R cannot run this code (takes forever)... the traceback says "no loop for break/next, jumping to top level", even though I do not have any break or next statements... I'm confused

I tried to divide the code into smaller code snippets, creating a data frame that is with r1 data and another with r2 data, to then merge them

```{r splitting code from attempt 1 to combine replicates}
mcl_testdata_r1 <- mcl_testdata %>%
  dplyr::select(uniprot_id, contains("r1")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "Prot_id1",
                        values_to = "count_R1")

mcl_testdata_r2 <- mcl_testdata %>%
    dplyr::select(uniprot_id, contains("r2")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "Prot_id2",
                        values_to = "count_R2")

head(mcl_testdata_r1, n = 12)
head(mcl_testdata_r2, n = 12)
```

I want to merge & create a new column for the mean of the 2 replicates, ignoring NAs.

```{r attempt 2}
mcl_testdata_long <- cbind(mcl_testdata_r1, mcl_testdata_r2[, -1]) # [, -1] was to leave out the second uniprot_id column

mcl_testdata_avg <- mcl_testdata_long %>%
  rowwise() %>%
  mutate(avg_count = mean(c(count_R1, count_R2), na.rm = TRUE)) 

#reformat the NaN in the new column to NA
mcl_testdata_avg[mcl_testdata_avg == "NaN"] <- NA_integer_
```

### Visualize NAs

```{r visualize NAs}
comb1 <- mcl_testdata_avg %>%
  group_by(Prot_id1) %>%
  summarise(na_rate = sum(is.na(avg_count))) %>%
  mutate(Prot_id2 = Prot_id1) %>% # duplicates Prot_id1 into a new column Prot_id2
  separate(Prot_id2, c("remove", "Plex"), sep = "r1_") %>% # separates values in Prot_id2 into 2 parts, the part before "r1_" is assigned to new col "remove", the part afte "r1_" is assigned to "Plex"
  mutate(na_perc = na_rate / nrow(mcl_testdata)) # percentage of NA for each group

ggplot(comb1, aes(
    x = forcats::fct_reorder(Prot_id1, na_perc),
    y = na_perc,
    fill = Plex
  )) +
  geom_col() +
  theme_cowplot() +
  labs(x = "Prot_id", y = "[%] of proteins missing") +
  theme(
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.y =  element_line(colour = "black"),
    axis.line.x =  element_line(colour = "black"),
    axis.text.y = element_text(colour = "black"),
    legend.title = element_blank(),
    axis.text.x = element_text(
      size = 6,
      angle = 90,
      hjust = 1,
      vjust = 1
    ),
    legend.position = "none"
  ) 
```

do the same with a different color palette?

```{r visualize NA with different color palette}
plex_vec <- unique(
  stringr::str_remove(
    colnames(mcl_testdata[,-c(1,2)]), 
    "_._r1_|_._r2_|_.._r1_|_.._r2_")
  ) 

palette <- distinctColorPalette(length(plex_vec))

comb <- mcl_testdata_avg %>% 
  dplyr::filter(Prot_id1 %in% colnames(mcl_testdata[,-1])) %>%
  group_by(Prot_id1) %>%
  summarise(na_rate = sum(is.na(avg_count))) %>%
  mutate(Prot_id2 = Prot_id1) %>%
  separate(Prot_id2, c("remove", "Plex"), sep = "r1_") %>%
  mutate(na_perc = na_rate / nrow(mcl_testdata)) 

ggplot(comb, aes(
    x = forcats::fct_reorder(Prot_id1, na_perc),
    y = na_perc,
    fill = Plex
  )) +
  geom_col() +
  theme_cowplot() +
  labs(x = "Prot_id", y = "[%] of proteins missing") +
  theme(
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.y =  element_line(colour = "black"),
    axis.line.x =  element_line(colour = "black"),
    axis.text.y = element_text(colour = "black"),
    legend.title = element_blank(),
    axis.text.x = element_text(
      size = 5,
      angle = 90,
      hjust = 1,
      vjust = 1
    ),
    legend.position = "none"
  )  +
  scale_fill_manual(values = palette)
```

### Print metrics for missing values

```{r print metrics for NA}
print(paste("Total number of proteins identified:", nrow(mcl_testdata), sep = " "))
print(paste("Median number of proteins identified per sample:", nrow(mcl_testdata) - median(comb$na_rate), sep = " "))
print(paste("Minimum number of proteins identified per sample:", nrow(mcl_testdata) - max(comb$na_rate), sep = " "))
print(paste("Maximum number of proteins identified per sample:", nrow(mcl_testdata) - min(comb$na_rate), sep = " "))
```

## Replicate correlation

each plex was measured in two replicates - that should be similar - so we investigate the correlation between both replicates

```{r Heatmap}
library(ComplexHeatmap)
library(circlize)

R1 <- mcl_testdata %>%
  dplyr::select(contains("r1"))
R2 <- mcl_testdata %>%
  dplyr::select(contains("r2"))

#set the color-vector 
color_fun_corr = colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))

cor(R1, R2, method = "spearman", use = "pairwise.complete.obs") %>%
  Heatmap(as.matrix(.),
                   column_title = "Replicate 1",
                   row_title = "Replicate 2",
                   col = color_fun_corr,
                   show_row_names = FALSE,
                   show_column_names = FALSE,
                   cluster_rows = FALSE,
                   cluster_columns = FALSE,
                   name = "Spearman's R") %>%
  draw()
```


### Reference channel correlation

reference channel is the 11th sample, so I only select the 11th sample of each plex as ref_cor

Then, cor() calculates the pairwise Spearman correlations for the selected columns starting with 11, considering only complete observations (pairwise.complete.obs).
The resulting correlation matrix is converted to a data frame using as.data.frame().
The mutate() function adds new column "type" with the value "ref" to distinguish it as the reference-channel correlation.
pivot_longer() converts the correlation matrix from wide format to long format, resulting in a tibble with two columns: "Prot_id" and "count" (the correlation values) filter(count != 1) removes self-correlation values (correlation of a variable with itself), as these values are always 1 in correlation matrices.

```{r ref channel correlation}
mcl_testdata_wide <- mcl_testdata_avg %>%
  dplyr::select(c("uniprot_id", "Prot_id1", "avg_count")) %>%
  pivot_wider(names_from = "Prot_id1", values_from = avg_count)
#first, compute the reference-channel correlation 
ref_cor11 <- mcl_testdata_wide %>%
  dplyr::select(starts_with("11_")) %>% 
  cor(., method = "spearman", use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  mutate(type = "ref") %>%
  pivot_longer(!c(type), names_to = "Prot_id", values_to = "count") %>%
  filter(count != 1)

head(ref_cor11, n = 5)

ref_cor <- ref_cor11[duplicated(ref_cor11$count),]

head(ref_cor, n = 5)

ggplot(ref_cor, aes(count)) +
  geom_density(fill = "#2a9d8f") +
  cowplot::theme_cowplot() +
  labs(x = "Spearman's R" ) 
```

What I don't understand....
if the ref_cor11 has 10 columns after selecting the cols starting with 11 from mcl_testdata_wide - which does cor() choose to compare for correlation?
I don't really understand which count shows correlation between which samples in ref_cor and ref_cor11

Anyways, there seems to be a fair but widespread correlation for the reference channels.

## Raw data distribution

inspect how the data is distributed without any correction

```{r raw data distribution}
order_vec <- colnames(mcl_testdata[,-c(1:2)])

plot_grid(
  mcl_testdata %>%
    dplyr::select(-gene_id) %>%
    pivot_longer(!uniprot_id, names_to = "Prot_id", values_to = "intensity") %>%
    mutate(splitter = Prot_id) %>%
    separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
    filter(grepl("_r1_", Prot_id)) %>%
    ggplot(aes(
      factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
    )) +
    geom_boxplot() +
    theme_cowplot() +
    theme(axis.text.x = element_text(
      size = 6,
      angle = 90,
      hjust = 1,
      vjust = 1
    )) +
    labs(x = "", y = "log2(intensity)", title = "Replicate 1") +
    scale_fill_manual(values = palette),
  
  mcl_testdata %>%
    dplyr::select(-gene_id) %>%
    pivot_longer(!uniprot_id, names_to = "Prot_id", values_to = "intensity") %>%
    mutate(splitter = Prot_id) %>%
    separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
    filter(grepl("_r2_", Prot_id)) %>%
    ggplot(aes(
      factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
    )) +
    geom_boxplot() +
    theme_cowplot() +
    theme(axis.text.x = element_text(
      size = 6,
      angle = 90,
      hjust = 1,
      vjust = 1
    )) +
    labs(x = "", y = "log2(intensity)", title = "Replicate 2") +
    scale_fill_manual(values = palette),
  
  nrow = 2
)
```

## Sample loading normalization

In the first step, we normalize within each plex by creating a normalization factor from a global scaling value, here called "target" (median of all column sums) divided by each samples column sum.

```{r sample loading normalization}
library(dplyr)

## create a vector of plexes
plex_vec <- unique(
  stringr::str_remove(
    colnames(mcl_testdata[,-c(1,2)]), 
    "._r1_|._r2_|.._r1_|.._r2_")
  ) 

plex_vec
# Store each plex into a list embedment
plex_list <- list() 

for(i in plex_vec) {
  
  plex_list[[i]] <- mcl_testdata %>% 
    dplyr::select(contains(i))
  
}


#create the target scaling factor 
colsum_vec <- c() 
# ChatGPT says that colsum_vec <- numeric(length = total_number_of_columns) might be more efficient with large datasets

for(i in plex_vec) {
  
  df <- plex_list[[i]]
  
  colsum_vec <- c(colsum_vec, colSums(df, na.rm = TRUE))
  # colSums takes the sum of the whole column = sum of one plex
  
}

colsum_vec

target <- median(colsum_vec, na.rm = TRUE)

#correct per plex with a helper function
run_sl_correct <- function(Y) { 
  norm_facs <- target / colSums(Y, na.rm = TRUE) # claculate normalization factor for each column in matrix Y
  output_sl <- sweep(Y, 2, norm_facs, FUN = "*") # multiplies norm_facs with each column
  
  return(output_sl) # output_sl is the normalized matrix
}

data_sl_repl <- plex_list%>%
  purrr::map(run_sl_correct) %>% 
  bind_cols()

#Visualize the output 
plot_grid(
  data_sl_repl %>%
    tibble::rownames_to_column("n") %>% # R cannot find this function
    pivot_longer(!n, names_to = "Prot_id", values_to = "intensity") %>%
    mutate(splitter = Prot_id) %>%
    separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
    filter(grepl("_r1_", Prot_id)) %>%
    ggplot(aes(
      factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
    )) +
    geom_boxplot() +
    theme_cowplot() +
    theme(axis.text.x = element_text(
      size = 6,
      angle = 90,
      hjust = 1,
      vjust = 1
    )) +
    labs(x = "", y = "log2(intensity)", title = "Replicate 1") +
    scale_fill_manual(values = palette),
  
  data_sl_repl %>%
    rownames_to_column("n") %>% # R cannot find this function
    pivot_longer(!n, names_to = "Prot_id", values_to = "intensity") %>%
    mutate(splitter = Prot_id) %>%
    separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
    filter(grepl("_r2_", Prot_id)) %>%
    ggplot(aes(
      factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
    )) +
    geom_boxplot() +
    theme_cowplot() +
    theme(axis.text.x = element_text(
      size = 6,
      angle = 90,
      hjust = 1,
      vjust = 1
    )) +
    labs(x = "", y = "log2(intensity)", title = "Replicate 2") +
    scale_fill_manual(values = palette),
  
  nrow = 2
)
```

### Check for correlation between random sets of channels

```{r corr between random sets}
library(psych)

sl_test_data <- data_sl_repl %>%
  dplyr::select(contains("r1")) %>%
  dplyr::select(contains(plex_vec[1:6])) %>%
  dplyr::select(contains("10_"))

pairs.panels(log2(sl_test_data), lm = TRUE, main = "Random channel over plexes SL")
```

There is a fair amount of correlation, but still with room for improvement.
Let's further have a look into the general structure in the dataset by PCA.
Since we are working with a dataset that contains missing data, we probabilistic PCA (pPCA) that accounts for missing data <http://www.cs.columbia.edu/~blei/seminar/2020-representation/readings/TippingBishop1999.pdf>.

## Principal Component analysis

--\> highly correlated samples cluster together in a 2D grap The axes are ranked in order of importance - differences along x axis (PC1 = principle component 1) are more important than diff.
along the 2nd principal componenet on y-axis (PC2)

```{r pca}
library(pcaMethods)
library(scrime)

threshold <- 0.5

ppca_df <- data_sl_repl %>% 
  filter(rowMeans(is.na(.)) < threshold) %>%
  #feature-wise scaling and centering 
  rowScales() %>%
  as.data.frame()

data_ppca <- pcaMethods::pca(t(as.matrix(ppca_df)), method = "ppca", nPcs = 2, seed = 123)

ppca_out <- as.data.frame(scores(data_ppca)) %>%
  rownames_to_column("Prot_id") %>%
  separate(Prot_id, c("Number", "Plex"), sep = "_r1_|_r2_")
ggplot(ppca_out, aes(PC1, PC2, col = Plex)) +
  geom_point(size = 3) +
  scale_colour_manual(values = palette) +
  theme_classic() +
  xlab(paste("PC1 (", round(data_ppca@R2[1] * 100, digits = 2), "%)")) +
  ylab(paste("PC2 (", round(data_ppca@R2[2] * 100, digits = 2), "%)"))
```

The plexes cluster a lot, there are a few outliers (ex. in plex 1880).
Plex 1876 forms 2 clusters - perhaps the replicates are a little different for this plex?

## Internal reference scaling

```{r internal reference scaling}
#make a dataframe of the reference channels per plex 
irs_factors <- data_sl_repl %>% 
  dplyr::select(contains("11_"))
irs_factors

#calculate the geometric mean per sample 
irs_factors$geomean <- apply(irs_factors, 1, function(x) exp(mean(log(x), na.rm = TRUE)))

#pull out sample names to identify  !!! Adjust removal position to geomeam column
irs_factor_vec = as.vector(names(irs_factors[,-c((length(plex_vec)*2)+1)]))
irs_factor_vec

#create the protein and reference channel wise scaling factor
for(i in irs_factor_vec) {
  factor <-  irs_factors$geomean / (irs_factors %>% dplyr::select(i))
  irs_factors[,paste0("fac_",i)] <- factor
}

irs_mult <- colnames(irs_factors %>% dplyr::select(contains("fac"))) %>% as.data.frame

irs_mult <- cbind(irs_mult, irs_mult) 

colnames(irs_mult) <- c("factor", "exp")
irs_mult
######
irs_mult<- irs_mult %>%
  separate(exp, c("discard", "plex"), sep = "\\_11_") %>%
  dplyr::select(-discard) %>% 
  filter(plex !=  "r1_1874") # why this step?? i just took the first plex

irs_mult_vec = as.vector(irs_mult$plex)

irs_fac_filt <- irs_factors %>% dplyr::select(contains("fac"))

all_irs <- (data_sl_repl %>% dplyr::select(contains( "r1_1874"))) * unlist(irs_fac_filt %>% dplyr::select(contains( "r1_1874")))

for(i in irs_mult_vec) {
  all_irs <- cbind(all_irs, (data_sl_repl %>% dplyr::select(contains(i))) * unlist(irs_fac_filt %>% dplyr::select(contains(i))))
}

all_irs

#Visualize the output 
plot_grid(
  all_irs %>%
    rownames_to_column("n") %>%
    pivot_longer(!n, names_to = "Prot_id", values_to = "intensity") %>%
    mutate(splitter = Prot_id) %>%
    separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
    filter(grepl("_r1_", Prot_id)) %>%
    ggplot(aes(
      factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
    )) +
    geom_boxplot() +
    theme_cowplot() +
    theme(axis.text.x = element_text(
      size = 6,
      angle = 90,
      hjust = 1,
      vjust = 1
    )) +
    labs(x = "", y = "log2(intensity)", title = "Replicate 1") +
    scale_fill_manual(values = palette),
  
  all_irs %>%
    rownames_to_column("n") %>%
    pivot_longer(!n, names_to = "Prot_id", values_to = "intensity") %>%
    mutate(splitter = Prot_id) %>%
    separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
    filter(grepl("_r2_", Prot_id)) %>%
    ggplot(aes(
      factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
    )) +
    geom_boxplot() +
    theme_cowplot() +
    theme(axis.text.x = element_text(
      size = 6,
      angle = 90,
      hjust = 1,
      vjust = 1
    )) +
    labs(x = "", y = "log2(intensity)", title = "Replicate 2") +
    scale_fill_manual(values = palette),
  
  nrow = 2
)

```

looks pretty good

### Check for the correlation between the random set of channels

```{r correlation between random set of channels 2}
sl_test_data <- all_irs %>%
  dplyr::select(contains("r1")) %>%
  dplyr::select(contains(plex_vec[1:6])) %>%
  dplyr::select(contains("10_"))

pairs.panels(log2(sl_test_data), lm = TRUE, main = "Random channel over plexes SL")

```

The correlation looks well improved!!

```{r check ppca}
ppca_df <- all_irs %>% 
  filter(rowMeans(is.na(.)) < threshold) %>%
  #feature-wise scaling and centering 
  rowScales() %>%
  as.data.frame()
data_ppca <- pcaMethods::pca(t(as.matrix(ppca_df)), method = "ppca", nPcs = 2, seed = 123)
ppca_out <- as.data.frame(scores(data_ppca)) %>%
  rownames_to_column("Prot_id") %>%
  separate(Prot_id, c("Number", "Plex"), sep = "_r1_|_r2_")
ggplot(ppca_out, aes(PC1, PC2, col = Plex)) +
  geom_point(size = 3) +
  scale_colour_manual(values = palette) +
  theme_classic() +
  xlab(paste("PC1 (", round(data_ppca@R2[1] * 100, digits = 2), "%)")) +
  ylab(paste("PC2 (", round(data_ppca@R2[2] * 100, digits = 2), "%)"))

```

The individual samples are spead and the plexes do not cluster anymore! 
The replicates still fall together - just as they should.

## Merge replicates by average
```{r merge replicates, eval = FALSE}
mcl_testdata_norm <- cbind(mcl_testdata[,c(1)], all_irs) %>%
  dplyr::select(-contains("11_"))

colnames(mcl_testdata_norm)[1] <- "uniprot_id"

mcl_testdata_long_norm <- cbind(
  mcl_testdata_norm %>%
    dplyr::select(uniprot_id, contains("r1")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "Prot_id1",
                        values_to = "count_R1"),
  mcl_testdata_norm %>%
    dplyr::select(uniprot_id, contains("r2")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "Prot_id2",
                        values_to = "count_R2") %>%
    dplyr::select(-uniprot_id)
  
) %>%
  rowwise() %>%
  dplyr::mutate(avg_count = mean(c_across(cols = c(count_R1, count_R2)), na.rm = TRUE))
```

Just like before, this code to merge the r1 and r2 data cannot be run by R... (takes forever), so again, I am dividing the code into smaller code snippets, creating a data frame that is with r1 data and another with r2 data, to then merge them:

```{r merge replicates in smaller codes}
mcl_testdata_norm <- cbind(mcl_testdata[,c(1)], all_irs) %>%
  dplyr::select(-contains("11_"))

colnames(mcl_testdata_norm)[1] <- "uniprot_id"

# create the datasets for r1 and r2 first separately
mcl_testdata_long_norm_r1 <- mcl_testdata_norm %>%
    dplyr::select(uniprot_id, contains("r1")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "Prot_id1",
                        values_to = "count_R1")

mcl_testdata_long_norm_r2 <- mcl_testdata_norm %>%
    dplyr::select(uniprot_id, contains("r2")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "Prot_id2",
                        values_to = "count_R2") 

head(mcl_testdata_long_norm_r1, n = 3)
head(mcl_testdata_long_norm_r2, n = 3)

# now merge:
# leave out uniprot_id (first col) from r2 dataset

mcl_testdata_long_norm <- cbind(mcl_testdata_long_norm_r1, mcl_testdata_long_norm_r2[, -1]) 

mcl_testdata_long_norm <- mcl_testdata_long_norm %>%
  rowwise() %>%
  mutate(avg_count = mean(c(count_R1, count_R2), na.rm = TRUE))
```
#### Question:
What is the difference between: 
mean(c_across(cols = c(count_R1, count_R2))
and
mean(c(count_R1, count_R2) ?

In this case - it seems like both would yield the same result.
The first code halted the code execution so I had to change it to the latter version.

```{r reformat NA}
#reformat the NA 
mcl_testdata_long_norm$avg_count[mcl_testdata_long_norm$avg_count == "NaN"] <- NA_integer_

mcl_testdata_wide_norm <- mcl_testdata_long_norm %>% 
  dplyr::select(uniprot_id, Prot_id1, avg_count) %>%
  pivot_wider(names_from = "Prot_id1", values_from = "avg_count")

```

Assign the names to the samples 

```{r assign names to samples}
m_comb <- as.matrix(mcl_testdata_wide_norm[,-1])

rownames(m_comb) <- mcl_testdata_wide_norm$uniprot_id

# t() function is used to transpose matrices and data frames meaning: it swaps rows and cols

t_comb <- t(m_comb) %>% 
  as.data.frame() %>%
  rownames_to_column("Pre_id") %>%
  separate(Pre_id, c("number", "plex"), sep = "_r1_") %>%
  mutate(plex = paste0("P", as.character(plex))) %>%
  mutate(plex = paste0(as.character(plex), as.character(number))) %>%
  dplyr::select(-number) %>% 
  column_to_rownames("plex")

mcl_testdata_proteome <-
  t(t_comb) %>% 
  as.data.frame() %>% 
  rownames_to_column("uniprot_id")

uniprot_gene_mcl_testdata <- mcl_testdata %>% dplyr::select(uniprot_id, gene_id) 

mcl_testdata_proteome_final <-  left_join(uniprot_gene_mcl_testdata, mcl_testdata_proteome, by  = "uniprot_id") 

mcl_testdata_proteome_final[,3:ncol(mcl_testdata_proteome_final)]<- sapply(mcl_testdata_proteome_final[,3:ncol(mcl_testdata_proteome_final)], as.numeric)

mcl_testdata_proteome_final[,3:ncol(mcl_testdata_proteome_final)] <- log2(mcl_testdata_proteome_final[,3:ncol(mcl_testdata_proteome_final)])

head(mcl_testdata_proteome_final, n = 5)
```

## Final data investigation
first, visualize the distribution per sample
```{r final data visualization}
mcl_testdata_wide_norm[,-c(1)] %>%
    rownames_to_column("n") %>%
    pivot_longer(!n, names_to = "Prot_id", values_to = "intensity") %>%
    mutate(splitter = Prot_id) %>%
    separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
    filter(grepl("_r1_", Prot_id)) %>%
    ggplot(aes(
      factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
    )) +
    geom_boxplot() +
    theme_cowplot() +
    theme(axis.text.x = element_text(
      size = 6,
      angle = 90,
      hjust = 1,
      vjust = 1
    )) +
    labs(x = "", y = "log2(intensity)") +
    scale_fill_manual(values = palette)
```

```{r final data ppca}
ppca_df <- mcl_testdata_proteome_final[,-c(1,2)] %>% 
  filter(rowMeans(is.na(.)) < threshold) %>%
  #feature-wise scaling and centering 
  rowScales() %>%
  as.data.frame()

data_ppca <- pcaMethods::pca(t(as.matrix(ppca_df)), method = "ppca", nPcs = 2, seed = 123)

ppca_out <- as.data.frame(scores(data_ppca)) %>%
  rownames_to_column("Prot_id") %>%
  separate(Prot_id, into = c("Plex", "Number"), sep = 5)
# sep = 5 means separate at the 5th string character

ggplot(ppca_out, aes(PC1, PC2, col = Plex)) +
  geom_point(size = 3) +
  scale_colour_manual(values = palette) +
  theme_classic() +
  xlab(paste("PC1 (", round(data_ppca@R2[1] * 100, digits = 2), "%)")) +
  ylab(paste("PC2 (", round(data_ppca@R2[2] * 100, digits = 2), "%)"))
```

```{r final daata NA distribution}
comb_df <- comb %>%
  dplyr::select(Plex, na_perc) %>%
  group_by(Plex) %>%
  summarise(na = mean(na_perc)) %>%
  mutate(Plex = paste0("P", Plex, sep = ""))

ppca_out <- as.data.frame(scores(data_ppca)) %>%
  rownames_to_column("Prot_id") %>%
  separate(Prot_id, c("Plex", "Number"), sep = 5) %>%
  left_join(comb_df, 
            by = "Plex")

ggplot(ppca_out, aes(PC1, PC2, col = na)) +
  geom_point(size = 3) +
  scale_color_viridis_c()+
  theme_classic() +
  xlab(paste("PC1 (", round(data_ppca@R2[1] * 100, digits = 2), "%)")) +
  ylab(paste("PC2 (", round(data_ppca@R2[2] * 100, digits = 2), "%)")) +
  labs(color = "% missing")
```

```{r}

```

```{r}

```
